Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-04-04 src/tex/ms.tex
GW910521 text

GW190521 is the most extreme representative of a class of events with bimodal $\kappa$ posteriors (Table \ref{tab:bimodal_events_mass}).
This appears to arise out of a confluence of factors related to this event's high mass and support for precession: GW190521 being the most massive event in our set, its signal is only in-band for a very short time before merger, meaning that frequency-dependent birefringence can more easily resemble spin effects.
In particular, a precession cycle can be approximated by a monotonic amplitude modulation for a short time.
This effect gives rise to a similar degeneracy between $\kappa$ and $\chi_p$ as outlined for GW170818, wherein the prior volume favored configurations with $|\kappa| >0$.
However, in the case of GW190521, the polarization measurement at the detectors is not accurate enough to fully disambiguate between right- and left-handed states (as implied by the bimodal $\cos\iota$ posterior obtained assuming \ac{GR} in Fig.~\ref{fig:corner_GW190521}).
As a result, the GW190521 posterior does not favor $\kappa > 0$ exclusively, but supports both $\kappa > 0$ and $\kappa < 0$, leading to a bimodal posterior.

>>>>

GW190521 is the most extreme representative of a class of events with bimodal $\kappa$ posteriors (Table \ref{tab:bimodal_events_mass}).
Figure \ref{fig:corner_GW190521} shows that the two peaks of the $\kappa$ distribution arise from respective modes in the inclination (lower left corner): one solution corresponds to an intrinsically face-on source ($\cos\iota \approx 1$) with negative $\kappa$, and the other to a face-off source ($\cos\iota \approx -1$) with positive $\kappa$.
In the former, an initially right-handed signal is birefringently enhanced by $\kappa < 0$ as it propagates towards the detector; in the latter, an initially left-handed signal is enhanced by $\kappa > 0$.
The two outcomes (a right-handed or left-handed signal at the detector) match the two possibilities allowed also in the reference \ac{GR} run, which itself yields a bimodal $\cos\iota$ posterior.
In both cases, the overall amplification of the signal introduced by a nonzero $\kappa$ is balanced by a large luminosity distance ($d_L \gtrsim 5 \, {\rm Gpc}$), so as to match the observed signal-to-noise ratio.

Introducing birefringence affects our inference of the spins in the GW190521 system.
We illustrate this in Fig.~\ref{fig:corner_GW190521} through the posterior on the effective spin $\chi_{\rm eff}$, which is favored to take values closer to zero in the birefringent analysis;
this can also be seen clearly in the joint posterior for the individual spin magnitudes, provided in Appendix \ref{sec:corner_GW190521_appendix}.
This suggests that at least part of the data features that lead the \ac{GR} analysis to infer near extremal spins can be alternatively explained as a consequence of birefringence.
The fact that this is possible is not fully surprising given that the signal from this heavy system is only observed for a very small number of cycles \cite{LIGOScientific:2020iuh,LIGOScientific:2020ufj}.

All in all, the situation for GW190521 is not unlike that of GW170818 discussed above (Sec.~\ref{sec:GW170818}): features in the data that would normally be interpreted as traces of high component spins are instead (at least partially) attributed to birefringent propagation, driven by a prior preference for high distances.
The degeneracies in the case of GW190521 are likely exacerbated by the heavier mass of the system, which reduces the observed number of cycles.
Crucially, GW190521 also lacks a definite measurement of the polarization state at the detector: without certainty about the ratio of right- versus left-handed components in the observed signal, it is not possible to disambiguate between $\kappa > 0$ versus $\kappa < 0$; hence the $\kappa$ distribution remains bimodal for GW190521, unlike for GW170918.
The poor determination of the polarization state of the detected signal is evidenced by the bimodal $\cos\iota$ posterior inferred assuming \ac{GR} in Fig.~\ref{fig:corner_GW190521}, and is a consequence of several factors including sky location.
